---
title: "Montoro et al. structure plots"
author: "Jason Willwerscheid"
date: "2/2/2025"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

## Introduction

Here I produce structure plots for EBNMF, NMF, and topic model fits to the larger dataset from Montoro et al.

Load the Tidyverse:

```{r, message=FALSE}
library(tidyverse)
library(cowplot)
library(ggrepel)
library(reactable)
library(htmltools)
```

Load the `flashier` results:

```{r}
scale_FF <- function(FF, LL, D = 1) {
  LL_norms <- apply(LL, 2, function(x) sqrt(sum(x^2)))
  return(t(t(FF) * D * LL_norms))
}

ebnmf <- readRDS("output/montoro-ebmf.rds")
ebnmf_FF <- scale_FF(ebnmf$fit$F.pm, ebnmf$fit$L.pm)
```

Extract cell covariates:

```{r}
cell_types <- c(
      "Ciliated",
      "Proliferating",
      "Basal",
      "Club",
      "Club (hillock-associated)",
      "Goblet.progenitor",
      "Goblet.1",
      "Goblet.2",
      "Tuft.progenitor",
      "Tuft.1",
      "Tuft.2",
      "Neuroendocrine",
      "Ionocyte"
)

ct_abbr <- cell_types
ct_abbr <- str_replace(ct_abbr, "\\.", "\\-")
ct_abbr[2] <- "Prolif."
ct_abbr[5] <- "Hillock"
ct_abbr[12] <- "NEC"
ct_abbr <- str_replace(ct_abbr, "progenitor", "Pr.")

external_info <- tibble(
  CellType = sapply(strsplit(rownames(ebnmf_FF), "_"), `[[`, 5),
  TimePoint = sapply(strsplit(rownames(ebnmf_FF), "_"), `[[`, 2),
  Replicate = sapply(strsplit(rownames(ebnmf_FF), "_"), `[[`, 3)
) |>
  left_join(tibble(CellType = cell_types, CtAbbr = ct_abbr), by = "CellType") |>
  mutate(
    CellType = factor(CellType, levels = cell_types),
    CtAbbr = factor(CtAbbr, levels = ct_abbr),
    TimePoint = factor(TimePoint),
    Replicate = factor(Replicate),
  ) |>
  mutate(
    TPxRep = fct_cross(Replicate, TimePoint),
    CommonType = fct_collapse(CtAbbr, Other = ct_abbr[6:13])
  )
```

I will use the following functions to generate plots:

```{r}
make_tib <- function(FF, info_col, ksort, ncells = 50, duprate = 1) {
  info <- external_info[[info_col]]

  # Downsample the number of cells.
  set.seed(666)
  cell_idx <- numeric(0)
  for (k in levels(info)) {
    which_idx <- which(info == k)
    # Downsample common cell types. Duplicate rare ones.
    if (ncells < length(which_idx)) {
      which_idx <- sample(which_idx, ncells, replace = FALSE)
    } else if (ncells >= length(which_idx) * duprate) {
      which_idx <- rep(which_idx, duprate)
    }
    cell_idx <- c(cell_idx, which_idx)
  }

  FF <- FF[cell_idx, ]
  info <- info[cell_idx]
  colnames(FF) <- 1:ncol(FF)

  lvls <- as.numeric(ksort)
  if (any(setdiff(1:ncol(FF), ksort))) {
    lvls <- c("Other", lvls)
  }

  tib <- as_tibble(FF) |>
    mutate(
      CellIdx = row_number(),
      Info = info
    ) |>
    pivot_longer(
      -c(CellIdx, Info),
      names_to = "Component",
      values_to = "Loading",
      values_drop_na = TRUE
    ) |>
    mutate(
      Component = ifelse(as.numeric(Component) %in% ksort, Component, "Other")
    ) |>
    group_by(CellIdx, Info, Component) |>
    summarize(Loading = sum(Loading), .groups = "drop") |>
    mutate(
      Component = factor(Component, levels = lvls)
    ) |>
    arrange(Info, Component)
  
  levels(tib$Component) <- c("Other", names(ksort))

  return(tib)
}

do_plot <- function(tib, colors, ptitle, ntiles = 400) {
  struct_df <- tib |>
    pivot_wider(names_from = Component, values_from = Loading)
  Lmat <- struct_df |>
    select(-(1:2)) |>
    as.matrix()
  group <- struct_df$Info
  
  if ("Other" %in% tib$Component) {
    colors <- c("grey90", colors)
  }
  
  p <- fastTopics::structure_plot(
    Lmat, topics = colnames(Lmat),
    grouping = group,
    colors = colors,
    # embed_method = fastTopics::umap_from_topics,
    # dims = 1,
    gap = 10,
    verbose = FALSE
  ) + theme(
    axis.text.y = element_blank()
  ) + labs(
    y = "",
    title = ptitle
  )

  return(p)
}

glasbey <- function(k) {
  return(fastTopics:::glasbey()[k])
}
```

The following function is used to generate tables of gene set annotations:

```{r}
make_gene_set_tbl <- function(GSEA, ksort) {
  klabs <- tibble(
    korig = sort(ksort),
    ksort = ksort,
    label = factor(names(ksort), levels = names(ksort))
  )

  keep_res <- GSEA |>
    filter(enrichmentScore > 0) |>
    left_join(klabs, by = c("k" = "ksort")) |>
    group_by(k) |>
    filter(FDR < 0.05 | rank(FDR) == min(rank(FDR))) |>
    ungroup()

  shared_genesets <- keep_res |>
    left_join(keep_res |> select(geneSet, shared_by = label),
              by = "geneSet",
              relationship = "many-to-many") |>
    filter(shared_by != label) |>
    arrange(label) |>
    group_by(label, geneSet) |>
    summarize(shared_by = paste(shared_by, collapse = ", "))
  
  keep_res <- keep_res |>
    select(label, geneSet, description, FDR) |>
    left_join(shared_genesets, by = c("label", "geneSet")) |>
    arrange(label, FDR, description)
  
  keep_res <- keep_res |>
    mutate(FDR = ifelse(FDR < 0.01, "<.01", format(round(FDR, 2), nsmall = 2)))
  
  react_tbl <- reactable(
    keep_res, 
    columns = list(
      label = colDef(name = "Component"),
      geneSet = colDef(name = "Gene Set"),
      description = colDef(name = "Description"),
      shared_by = colDef(name = "Shared By")
    ),
    groupBy = "label"
  )
  
  return(react_tbl)
}
```

And the following is used to make "top gene" tables:

```{r}
bar_chart <- function(width = "100%", height = "1rem", fill = "#00bfc4", background = NULL) {
  bar <- div(style = list(background = fill, width = width, height = height))
  chart <- div(style = list(flexGrow = 1, marginLeft = "0.5rem", background = background), bar)
  div(style = list(display = "flex", alignItems = "center"), chart)
}

bar_chart_pos_neg <- function(label, value, max_value = 1, height = "1rem",
                              pos_fill = "#005ab5", neg_fill = "#dc3220") {
  neg_chart <- div(style = list(flex = "1 1 0"))
  pos_chart <- div(style = list(flex = "1 1 0"))
  width <- paste0(abs(value / max_value) * 100, "%")

  if (value < 0) {
    bar <- div(style = list(marginLeft = "0.5rem", background = neg_fill, width = width, height = height))
    chart <- div(
      style = list(display = "flex", alignItems = "center", justifyContent = "flex-end"),
      label,
      bar
    )
    neg_chart <- tagAppendChild(neg_chart, chart)
  } else {
    bar <- div(style = list(marginRight = "0.5rem", background = pos_fill, width = width, height = height))
    chart <- div(style = list(display = "flex", alignItems = "center"), bar, label)
    pos_chart <- tagAppendChild(pos_chart, chart)
  }

  div(style = list(display = "flex"), neg_chart, pos_chart)
}

make_top_genes_tbl <- function(LL, ksort, gene_names, top_n = 30) {
  LL <- t(t(LL) / apply(LL, 2, max)) # Scale so maximum loading for each k is 1
  LL <- LL[, ksort]
  colnames(LL) <- names(ksort)
  LL_tib <- as_tibble(LL) |>
    mutate(Gene = gene_names) |>
    pivot_longer(-Gene, names_to = "Component", values_to = "Loading") |>
    mutate(Component = factor(Component, levels = names(ksort))) |>
    group_by(Gene) |>
    mutate(Uniqueness = log10(Loading) - log10(sort(Loading, decreasing = TRUE)[2]))
  
  shared_genes <- LL_tib |>
    left_join(LL_tib |> select(Gene, OtherLoading = Loading, SharedBy = Component),
              by = join_by(Gene == Gene, Loading < OtherLoading),
              relationship = "many-to-many") |>
    filter(SharedBy != Component) |>
    arrange(Component) |>
    group_by(Component, Gene) |>
    summarize(SharedBy = paste(SharedBy, collapse = ", "))
  
  top_genes <- LL_tib |>
    group_by(Component) |>
    slice_max(Loading, n = top_n) |>
    left_join(shared_genes, by = c("Component", "Gene")) |>
    ungroup()

  react_tbl <- reactable(
    top_genes, 
    columns = list(
      Loading = colDef(name = "Loading", align = "right", cell = function(value) {
        width <- paste0(value * 100, "%")
        bar_chart(width = width, background = "#e1e1e1")
      }),
      Uniqueness = colDef(name = "Uniqueness", align = "center", cell = function(value) {
        label <- round(value, 2)
        bar_chart_pos_neg(label, value)
      }),
      SharedBy = colDef(name = "Larger In")
    ),
    groupBy = "Component"
  )

  return(react_tbl)
}
```



## EBNMF

In each case I manually sort and label factors.

```{r fig.height=12, fig.width=8, message=FALSE}
ebnmf_ksort <- c(
  26, 27, 30, # Sparse
  1, 8, 10, 16, # Dense
  11, 12, 2, 24, # Time points
  6, 13, 22, # Ciliated
  21, 23, 17, # Proliferating
  4, 5, 9, # Basal
  3, 15, 7, # 26, Club-Hillock
  25, 28, 29, # Goblet
  18, 19, 14, 20 # Tuft-NEC-Ionocyte
)
names(ebnmf_ksort) <- c(
  "Sprs-a", "Sprs-b", "Sprs-c",
  "Dns-a", "Dns-b", "Dns-c", "Dns-d",
  "Tp0", "Tp30", "Tp60", "R1xTp30",
  "Cil-a", "Cil-b", "Cil-c",
  "Prl-a", "Prl-b", "Prl-c",
  "Bas-a", "Bas-b", "Bas-c",
  "Cl/G-a", "Cl/G-b", "Hil",
  "Gob", "Gob1", "Gob2",
  "Tuf1", "Tuf2", "Nec", "Ion"
)

sparse_title <- "Very sparse factors"
dense_title <- "Dense factors"
tp_title <- "Time point factors"
common_title <- "Prevalent cell type factors"
rare_title <- "Rare cell type factors (common cell types are downsampled)"

sparse_k <- 1:3
ebnmf_tib <- make_tib(ebnmf_FF, "CommonType", ebnmf_ksort[sparse_k], Inf, 1)
sparse_p <- do_plot(ebnmf_tib, glasbey(2:4), sparse_title)

dense_k <- 4:7
ebnmf_tib <- make_tib(ebnmf_FF, "CommonType", ebnmf_ksort[dense_k], Inf, 1)
dense_p <- do_plot(ebnmf_tib, glasbey(5:8), dense_title)

tp_k <- 8:11
ebnmf_tib <- make_tib(ebnmf_FF, "TPxRep", ebnmf_ksort[tp_k], Inf, 1)
tp_p <- do_plot(ebnmf_tib, glasbey(14:17), tp_title)

common_k <- c(12:14, 18:20, 23)
ebnmf_tib <- make_tib(ebnmf_FF, "CommonType", ebnmf_ksort[common_k], Inf, 1)
common_p <- do_plot(ebnmf_tib, glasbey(21:27), common_title)

rare_k <- c(15:17, 21:22, 24:30)
ebnmf_tib <- make_tib(ebnmf_FF, "CtAbbr", ebnmf_ksort[rare_k], 100, 3)
rare_p <- do_plot(ebnmf_tib, glasbey(32:43), rare_title)

plot_grid(common_p, rare_p, tp_p, dense_p, sparse_p,  nrow = 5, ncol = 1)
```

## NMF

```{r fig.height=10.5, fig.width=8, message=FALSE}
nnlm <- readRDS("output/montoro-nmf.rds")
nnlm_FF <- scale_FF(t(nnlm$fit$h), nnlm$fit$w, nnlm$fit$d)

nnlm_ksort <- c(
  1, 28, 10, 26, # dense
  9, 5, # time points (dense)
  3, 11, 8, 7, # time points
  21, # replicate x TP
  15, 22, 20, # ciliated
  24, 12, 19, # prolif
  2, 6, 4, 14, 13, # basal
  25, 17, 23, # club/goblet
  18, 16, # hillock
  29, # tuft
  27, # NEC
  30 # ionocyte
)
names(nnlm_ksort) <- c(
  "Dns-a", "Dns-b", "Dns-c", "Dns-d",
  "TpxCl", "Tp30/60",
  "Tp0-a", "Tp0-b", "Tp30", "Tp60",
  "R1xTp30",
  "Cil-a", "Cil-b", "Cil-c",  
  "Prl-a", "Prl-b", "Prl-c",
  "Bas-a", "Bas-b", "Bas-c", "Bas-d", "Bas-e",
  "Cl/G-a", "Cl/G-b", "Cl/G-c",
  "Hil-a", "Hil-b",
  "Tuf",
  "Nec",
  "Ion/Cl/G"
)

nnlm_dense_k <- 1:4
nnlm_tib <- make_tib(nnlm_FF, "CommonType", nnlm_ksort[nnlm_dense_k], Inf, 1)
nnlm_dense_p <- do_plot(nnlm_tib, glasbey(5:8), dense_title)

nnlm_tp_k <- 5:11
nnlm_tib <- make_tib(nnlm_FF, "TPxRep", nnlm_ksort[nnlm_tp_k], Inf, 1)
nnlm_tp_p <- do_plot(nnlm_tib, glasbey(11:17), tp_title)

nnlm_common_k <- c(12:14, 18:22, 26:27)
nnlm_tib <- make_tib(nnlm_FF, "CommonType", nnlm_ksort[nnlm_common_k], Inf, 1)
nnlm_common_p <- do_plot(nnlm_tib, glasbey(21:30), common_title)

nnlm_rare_k <- c(15:17, 23:25, 28:30)
nnlm_tib <- make_tib(nnlm_FF, "CtAbbr", nnlm_ksort[nnlm_rare_k], 100, 3)
nnlm_rare_p <- do_plot(nnlm_tib, glasbey(32:40), rare_title)

plot_grid(nnlm_common_p, nnlm_rare_p, nnlm_tp_p, nnlm_dense_p, nrow = 4, ncol = 1)
```

## Topic model

```{r fig.height=10.5, fig.width=8, message=FALSE}
tm <- readRDS("output/montoro-topics.rds")
tm_fit <- tm$fit
tm_fit$F <- tm$fit$L
tm_fit$L <- tm$fit$F
tm_fit$Fn <- tm$fit$Ln
tm_fit$Ln <- tm$fit$Fn
tm_fit$Fy <- tm$fit$Ly
tm_fit$Ly <- tm$fit$Fy
tm_fit <- fastTopics::poisson2multinom(tm_fit)
tm_FF <- tm_fit$L

tm_ksort <- c(
  21, 5, 11, 17, 27, # Dense  
  19, # Tp30/60xCl
  15, 8, # Tp0/60?
  24, # Tp30/60
  14, 26, # Tp0
  23, 29, 9, # Tp30 - 23 also club
  13, # Tp60
  25, # Ciliated
  6, # Proliferating
  10, 4, # Basal/Prolif
  12, # Basal
  3, # Club/Goblet
  18, # Club/Hillock
  22, # Club
  16, 28, # Hillock/Goblet-progenitor
  20, # Hillock
  7, # Goblet-p/1
  2, # Goblet-2
  1, 30 # Tuft-NEC-Ion
)
names(tm_ksort) <- c(
  "Dns-a", "Dns-b", "Dns-c",
  "Dns-d", "Dns-e",
  "TpxCl",
  "Tp0/60-a", "Tp0/60-b",
  "Tp30/60",
  "Tp0-a", "Tp0-b",
  "Tp30xCl", "Tp30-a", "Tp30-b",
  "Tp60",
  "Cil",
  "Prl",
  "B/P-a", "B/P-b",
  "Bas", 
  "Cl/G", "Cl/Hil", "Club",
  "H/G-a", "H/G-b", 
  "Hil", 
  "Gob1", "Gob2",
  "TNI-a", "TNI-b"
)

tm_dense_k <- 1:5
tm_tib <- make_tib(tm_FF, "CommonType", tm_ksort[tm_dense_k], Inf, 1)
tm_dense_p <- do_plot(tm_tib, glasbey(5:9), dense_title)

tm_tp_k <- 6:15
tm_tib <- make_tib(tm_FF, "TPxRep", tm_ksort[tm_tp_k], Inf, 1)
tm_tp_p <- do_plot(tm_tib, glasbey(11:20), tp_title)

tm_common_k <- c(16, 20, 22:23, 26)
tm_tib <- make_tib(tm_FF, "CommonType", tm_ksort[tm_common_k], Inf, 1)
tm_common_p <- do_plot(tm_tib, glasbey(24:29), common_title)

tm_rare_k <- c(17:19, 21, 24:25, 27:30)
tm_tib <- make_tib(tm_FF, "CtAbbr", tm_ksort[tm_rare_k], 100, 3)
tm_rare_p <- do_plot(tm_tib, glasbey(32:41), rare_title)

plot_grid(tm_common_p, tm_rare_p, tm_tp_p, tm_dense_p, nrow = 4, ncol = 1)
```

## EBNMF annotations and top genes

The first table provides gene set annotations. I use GSEA. For each component, I provide as input all genes with loadings that are larger than 0.01 times the maximum loading (for the topic model, 0.001 times the maximum loading). I report the top ten gene sets with FDR < 0.05 for each component (ties are included), with the restriction that at least one gene set is reported for each factor (even if its FDR is greater than 0.05).

The second table lists the 30 largest gene loadings for each component. "Uniqueness" is measured as the log ratio of the gene loading to the second-largest loading across all (l1-normalized) components. If a gene loading is large in only one component, its "uniqueness" will be positive and large for that component (positive values are shown as blue bars). If it is larger in exactly one component, its uniqueness will be equal to 0. If it is larger in more than one component, its uniqueness will be negative (shown as a red bar).

```{r message=FALSE}
ebnmf_GSEA <- readRDS("output/montoro-ebnmf-gsea.rds")
make_gene_set_tbl(ebnmf_GSEA, ebnmf_ksort[c(12:30, 8:11, 4:7, 1:3)])
make_top_genes_tbl(ebnmf$fit$L.pm, ebnmf_ksort[c(12:30, 8:11, 4:7, 1:3)], rownames(ebnmf$fit$L.pm))
```

## NMF annotations and top genes

```{r message=FALSE}
nmf_GSEA <- readRDS("output/montoro-nmf-gsea.rds")
make_gene_set_tbl(nmf_GSEA, nnlm_ksort[c(12:30, 5:11, 1:4)])
make_top_genes_tbl(nnlm$fit$w, nnlm_ksort[c(12:30, 5:11, 1:4)], rownames(ebnmf$fit$L.pm))
```

## Topic model annotations and top genes

```{r message=FALSE}
tm_GSEA <- readRDS("output/montoro-tm-gsea.rds")
make_gene_set_tbl(tm_GSEA, tm_ksort[c(16:30, 6:15, 1:5)])
make_top_genes_tbl(tm_fit$F, tm_ksort[c(16:30, 6:15, 1:5)], rownames(ebnmf$fit$L.pm))
```

## Tuft-1 vs Tuft-2

EBNMF is the only method that picks up the difference between Tuft-1 and Tuft-2 cells. However, the structure plots suggest that the difference might better be understood as a gradient between two extremes rather than a binary distinction.

Montoro et al. characterize the Tuft-1 subset as more strongly expressing genes associated with taste transduction, while the Tuft-2 subset is characterized as more strongly expressing inflammatory genes (see Montoro et al. (2018), Figure 4). 

The top "unique" genes listed above for each component are similar to the genes highlighted in the paper: Gnb3, Gng13, Fxyd6, and Ovol3 for Tuft-1; and Alox5ap and Dclk1 for Tuft-2. A more direct comparison of loadings in the "Tuf1" and "Tuf2" components picks up other highlighted genes, including Gnat3 for Tuft-1 and Mgst3 for Tuft-2:

```{r}
tuft_LL <- ebnmf$fit$L.pm[, ebnmf_ksort[c("Tuf1", "Tuf2")]]
tuft_LL <- t(t(tuft_LL) / apply(tuft_LL, 2, max))
tuft_tib <- tibble(
  Gene = rownames(tuft_LL),
  Tuf1 = tuft_LL[, 1],
  Tuf2 = tuft_LL[, 2],
  Label = ifelse(abs(Tuf1 - Tuf2) > 0.22, Gene, "")
) 
ggplot(tuft_tib, aes(Tuf1, Tuf2)) +
  theme_minimal() +
  geom_point(size = 0.5) +
  geom_text_repel(aes(label = Label), size = 2, min.segment.length = 0) +
  scale_fill_gradient(trans = "log10")
```
