---
title: "Montoro et al. structure plots"
author: "Jason Willwerscheid"
date: "2/2/2025"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

## Introduction

Here I produce structure plots for EBNMF, NMF, and topic model fits to the larger dataset from Montoro et al.

Load the Tidyverse:

```{r, message=FALSE}
library(tidyverse)
library(cowplot)
library(gt)
```

Load the `flashier` results:

```{r}
scale_FF <- function(FF, LL, D = 1) {
  LL_norms <- apply(LL, 2, function(x) sqrt(sum(x^2)))
  return(t(t(FF) * D * LL_norms))
}

ebnmf <- readRDS("output/montoro-ebmf.rds")
ebnmf_FF <- scale_FF(ebnmf$fit$F.pm, ebnmf$fit$L.pm)
```

Extract cell covariates:

```{r}
cell_types <- c(
      "Ciliated",
      "Proliferating",
      "Basal",
      "Club",
      "Club (hillock-associated)",
      "Goblet.progenitor",
      "Goblet.1",
      "Goblet.2",
      "Tuft.progenitor",
      "Tuft.1",
      "Tuft.2",
      "Neuroendocrine",
      "Ionocyte"
)

ct_abbr <- cell_types
ct_abbr <- str_replace(ct_abbr, "\\.", "\\-")
ct_abbr[2] <- "Prolif."
ct_abbr[5] <- "Hillock"
ct_abbr[12] <- "NEC"
ct_abbr <- str_replace(ct_abbr, "progenitor", "Pr.")

external_info <- tibble(
  CellType = sapply(strsplit(rownames(ebnmf_FF), "_"), `[[`, 5),
  TimePoint = sapply(strsplit(rownames(ebnmf_FF), "_"), `[[`, 2),
  Replicate = sapply(strsplit(rownames(ebnmf_FF), "_"), `[[`, 3)
) |>
  left_join(tibble(CellType = cell_types, CtAbbr = ct_abbr), by = "CellType") |>
  mutate(
    CellType = factor(CellType, levels = cell_types),
    CtAbbr = factor(CtAbbr, levels = ct_abbr),
    TimePoint = factor(TimePoint),
    Replicate = factor(Replicate),
  ) |>
  mutate(
    TPxRep = fct_cross(Replicate, TimePoint),
    CommonType = fct_collapse(CtAbbr, Other = ct_abbr[6:13])
  )
```

I will use the following functions to generate plots:

```{r}
make_tib <- function(FF, info_col, ksort, ncells = 50, duprate = 1) {
  info <- external_info[[info_col]]

  # Downsample the number of cells.
  set.seed(666)
  cell_idx <- numeric(0)
  for (k in levels(info)) {
    which_idx <- which(info == k)
    # Downsample common cell types. Duplicate rare ones.
    if (ncells < length(which_idx)) {
      which_idx <- sample(which_idx, ncells, replace = FALSE)
    } else if (ncells >= length(which_idx) * duprate) {
      which_idx <- rep(which_idx, duprate)
    }
    cell_idx <- c(cell_idx, which_idx)
  }

  FF <- FF[cell_idx, ]
  info <- info[cell_idx]
  colnames(FF) <- 1:ncol(FF)

  lvls <- as.numeric(ksort)
  if (any(setdiff(1:ncol(FF), ksort))) {
    lvls <- c("Other", lvls)
  }

  tib <- as_tibble(FF) |>
    mutate(
      CellIdx = row_number(),
      Info = info
    ) |>
    pivot_longer(
      -c(CellIdx, Info),
      names_to = "Component",
      values_to = "Loading",
      values_drop_na = TRUE
    ) |>
    mutate(
      Component = ifelse(as.numeric(Component) %in% ksort, Component, "Other")
    ) |>
    group_by(CellIdx, Info, Component) |>
    summarize(Loading = sum(Loading), .groups = "drop") |>
    mutate(
      Component = factor(Component, levels = lvls)
    ) |>
    arrange(Info, Component)
  
  levels(tib$Component) <- c("Other", names(ksort))

  return(tib)
}

do_plot <- function(tib, colors, ptitle, ntiles = 400) {
  struct_df <- tib |>
    pivot_wider(names_from = Component, values_from = Loading)
  Lmat <- struct_df |>
    select(-(1:2)) |>
    as.matrix()
  group <- struct_df$Info
  
  if ("Other" %in% tib$Component) {
    colors <- c("grey90", colors)
  }
  
  p <- fastTopics::structure_plot(
    Lmat, topics = colnames(Lmat),
    grouping = group,
    colors = colors,
    # embed_method = fastTopics::umap_from_topics,
    # dims = 1,
    gap = 10,
    verbose = FALSE
  ) + theme(
    axis.text.y = element_blank()
  ) + labs(
    y = "",
    title = ptitle
  )

  return(p)
}

glasbey <- function(k) {
  return(fastTopics:::glasbey()[k])
}
```

The following function is used to generate tables of gene set annotations:

```{r}
make_tbl <- function(GSEA, ksort) {
  klabs <- tibble(
    korig = sort(ksort),
    ksort = ksort,
    label = factor(names(ksort), levels = names(ksort))
  )

  keep_res <- GSEA |>
    # Remove "identical" gene sets (userId lists genes found in set)
    left_join(klabs, by = c("k" = "ksort")) |>
    group_by(k, userId) |>
    arrange(FDR) |>
    slice_head(n = 1) |>
    group_by(k) |>
    slice_min(FDR, n = 5) |>
    filter(FDR < 0.05 | rank(FDR) == min(rank(FDR))) |>
    ungroup()

  shared_genesets <- keep_res |>
    group_by(geneSet) |>
    filter(n() > 1) |>
    ungroup()
  distinctive_genesets <- keep_res |>
    anti_join(shared_genesets)

  shared_genesets <- shared_genesets |>
    group_by(geneSet, description) |>
    arrange(label) |>
    summarize(components = paste0(label, collapse = ", "),
              FDR = ifelse(
                n() < 4,
                paste(ifelse(FDR < .01, "<.01", round(FDR, 2)), collapse = ", "),
                paste("all <", round(max(FDR) + 0.005001, 2))
              ),
              order1 = label[1], order2 = label[2],
              .groups = "drop") |>
    arrange(order1, order2) |>
    select(-order1, -order2)
  distinctive_genesets <- distinctive_genesets |>
    group_by(geneSet, description) |>
    summarize(components = paste0(label, collapse = ", "),
              order2 = FDR[1],
              FDR = paste(ifelse(FDR < .01, "<.01", round(FDR, 2)), collapse = ", "),
              order1 = label[1],
              .groups = "drop") |>
    arrange(order1, order2) |>
    select(-order1, -order2)

  genesets <- shared_genesets |>
    bind_rows(distinctive_genesets) |>
    group_by(components)

  gt_tbl <- gt(genesets) |>
    #cols_move_to_start(columns = components) |>
    cols_label(
      geneSet = "Gene Set",
      description = "Description",
      components = "Components"
    ) |>
    tab_style(
      style = cell_text(weight = "bold"),
      locations = cells_row_groups()
    ) |>
    tab_options(
      table.font.size = 11
    )
  return(gt_tbl)
}
```


## EBNMF

In each case I manually sort and label factors.

```{r fig.height=12, fig.width=8, message=FALSE}
ebnmf_ksort <- c(
  26, 27, 30, # Sparse
  1, 8, 10, 16, # Dense
  11, 12, 2, 24, # Time points
  6, 13, 22, # Ciliated
  21, 23, 17, # Proliferating
  4, 5, 9, # Basal
  3, 15, 7, # 26, Club-Hillock
  25, 28, 29, # Goblet
  18, 19, 14, 20 # Tuft-NEC-Ionocyte
)
names(ebnmf_ksort) <- c(
  "Sprs-a", "Sprs-b", "Sprs-c",
  "Dns-a", "Dns-b", "Dns-c", "Dns-d",
  "Tp0", "Tp30", "Tp60", "R1xTp30",
  "Cil-a", "Cil-b", "Cil-c",
  "Prl-a", "Prl-b", "Prl-c",
  "Bas-a", "Bas-b", "Bas-c",
  "Cl/G-a", "Cl/G-b", "Hil",
  "Gob", "Gob1", "Gob2",
  "Tuf1", "Tuf2", "Nec", "Ion"
)

sparse_title <- "Very sparse factors"
dense_title <- "Dense factors"
tp_title <- "Time point factors"
common_title <- "Prevalent cell type factors"
rare_title <- "Rare cell type factors (common cell types are downsampled)"

sparse_k <- 1:3
ebnmf_tib <- make_tib(ebnmf_FF, "CommonType", ebnmf_ksort[sparse_k], Inf, 1)
sparse_p <- do_plot(ebnmf_tib, glasbey(2:4), sparse_title)

dense_k <- 4:7
ebnmf_tib <- make_tib(ebnmf_FF, "CommonType", ebnmf_ksort[dense_k], Inf, 1)
dense_p <- do_plot(ebnmf_tib, glasbey(5:8), dense_title)

tp_k <- 8:11
ebnmf_tib <- make_tib(ebnmf_FF, "TPxRep", ebnmf_ksort[tp_k], Inf, 1)
tp_p <- do_plot(ebnmf_tib, glasbey(14:17), tp_title)

common_k <- c(12:14, 18:20, 23)
ebnmf_tib <- make_tib(ebnmf_FF, "CommonType", ebnmf_ksort[common_k], Inf, 1)
common_p <- do_plot(ebnmf_tib, glasbey(21:27), common_title)

rare_k <- c(15:17, 21:22, 24:30)
ebnmf_tib <- make_tib(ebnmf_FF, "CtAbbr", ebnmf_ksort[rare_k], 100, 3)
rare_p <- do_plot(ebnmf_tib, glasbey(32:43), rare_title)

plot_grid(common_p, rare_p, tp_p, dense_p, sparse_p,  nrow = 5, ncol = 1)
```

## NMF

```{r fig.height=10.5, fig.width=8, message=FALSE}
nnlm <- readRDS("output/montoro-nmf.rds")
nnlm_FF <- scale_FF(t(nnlm$fit$h), nnlm$fit$w, nnlm$fit$d)

nnlm_ksort <- c(
  1, 28, 10, 26, # dense
  9, 5, # time points (dense)
  3, 11, 8, 7, # time points
  21, # replicate x TP
  15, 22, 20, # ciliated
  24, 12, 19, # prolif
  2, 6, 4, 14, 13, # basal
  25, 17, 23, # club/goblet
  18, 16, # hillock
  29, # tuft
  27, # NEC
  30 # ionocyte
)
names(nnlm_ksort) <- c(
  "Dns-a", "Dns-b", "Dns-c", "Dns-d",
  "TpxCl", "Tp30/60",
  "Tp0-a", "Tp0-b", "Tp30", "Tp60",
  "R1xTp30",
  "Cil-a", "Cil-b", "Cil-c",  
  "Prl-a", "Prl-b", "Prl-c",
  "Bas-a", "Bas-b", "Bas-c", "Bas-d", "Bas-e",
  "Cl/G-a", "Cl/G-b", "Cl/G-c",
  "Hil-a", "Hil-b",
  "Tuf",
  "Nec",
  "Ion/Cl/G"
)

nnlm_dense_k <- 1:4
nnlm_tib <- make_tib(nnlm_FF, "CommonType", nnlm_ksort[nnlm_dense_k], Inf, 1)
nnlm_dense_p <- do_plot(nnlm_tib, glasbey(5:8), dense_title)

nnlm_tp_k <- 5:11
nnlm_tib <- make_tib(nnlm_FF, "TPxRep", nnlm_ksort[nnlm_tp_k], Inf, 1)
nnlm_tp_p <- do_plot(nnlm_tib, glasbey(11:17), tp_title)

nnlm_common_k <- c(12:14, 18:22, 26:27)
nnlm_tib <- make_tib(nnlm_FF, "CommonType", nnlm_ksort[nnlm_common_k], Inf, 1)
nnlm_common_p <- do_plot(nnlm_tib, glasbey(21:30), common_title)

nnlm_rare_k <- c(15:17, 23:25, 28:30)
nnlm_tib <- make_tib(nnlm_FF, "CtAbbr", nnlm_ksort[nnlm_rare_k], 100, 3)
nnlm_rare_p <- do_plot(nnlm_tib, glasbey(32:40), rare_title)

plot_grid(nnlm_common_p, nnlm_rare_p, nnlm_tp_p, nnlm_dense_p, nrow = 4, ncol = 1)
```

## Topic model

```{r fig.height=10.5, fig.width=8, message=FALSE}
tm <- readRDS("output/montoro-topics.rds")
tm_fit <- tm$fit
tm_fit$F <- tm$fit$L
tm_fit$L <- tm$fit$F
tm_fit$Fn <- tm$fit$Ln
tm_fit$Ln <- tm$fit$Fn
tm_fit$Fy <- tm$fit$Ly
tm_fit$Ly <- tm$fit$Fy
tm_fit <- fastTopics::poisson2multinom(tm_fit)
tm_FF <- tm_fit$L

tm_ksort <- c(
  21, 5, 11, 17, 27, # Dense  
  19, # Tp30/60xCl
  15, 8, # Tp0/60?
  24, # Tp30/60
  14, 26, # Tp0
  23, 29, 9, # Tp30 - 23 also club
  13, # Tp60
  25, # Ciliated
  6, # Proliferating
  10, 4, # Basal/Prolif
  12, # Basal
  3, # Club/Goblet
  18, # Club/Hillock
  22, # Club
  16, 28, # Hillock/Goblet-progenitor
  20, # Hillock
  7, # Goblet-p/1
  2, # Goblet-2
  1, 30 # Tuft-NEC-Ion
)
names(tm_ksort) <- c(
  "Dns-a", "Dns-b", "Dns-c",
  "Dns-d", "Dns-e",
  "TpxCl",
  "Tp0/60-a", "Tp0/60-b",
  "Tp30/60",
  "Tp0-a", "Tp0-b",
  "Tp30xCl", "Tp30-a", "Tp30-b",
  "Tp60",
  "Cil",
  "Prl",
  "B/P-a", "B/P-b",
  "Bas", 
  "Cl/G", "Cl/Hil", "Club",
  "H/G-a", "H/G-b", 
  "Hil", 
  "Gob1", "Gob2",
  "TNI-a", "TNI-b"
)

tm_dense_k <- 1:5
tm_tib <- make_tib(tm_FF, "CommonType", tm_ksort[tm_dense_k], Inf, 1)
tm_dense_p <- do_plot(tm_tib, glasbey(5:9), dense_title)

tm_tp_k <- 6:15
tm_tib <- make_tib(tm_FF, "TPxRep", tm_ksort[tm_tp_k], Inf, 1)
tm_tp_p <- do_plot(tm_tib, glasbey(11:20), tp_title)

tm_common_k <- c(16, 20, 22:23, 26)
tm_tib <- make_tib(tm_FF, "CommonType", tm_ksort[tm_common_k], Inf, 1)
tm_common_p <- do_plot(tm_tib, glasbey(24:29), common_title)

tm_rare_k <- c(17:19, 21, 24:25, 27:30)
tm_tib <- make_tib(tm_FF, "CtAbbr", tm_ksort[tm_rare_k], 100, 3)
tm_rare_p <- do_plot(tm_tib, glasbey(32:41), rare_title)

plot_grid(tm_common_p, tm_rare_p, tm_tp_p, tm_dense_p, nrow = 4, ncol = 1)
```

## EBNMF annotations

I use GSEA to do annotations. For each component, I provide as input all genes with loadings that are larger than 0.01 times the maximum loading (for the topic model, 0.001 times the maximum loading). (If all loadings are provided, GSEA will also find gene sets with very small loadings across genes, which is obviously not what we want.) I report the top 5 gene sets with FDR < 0.05 for each component (ties are included), with the restriction that at least one gene set is reported for each factor (even if its FDR is greater than 0.05).

```{r message=FALSE}
ebnmf_GSEA <- readRDS("output/montoro-ebnmf-gsea.rds")
make_tbl(ebnmf_GSEA, ebnmf_ksort[c(12:30, 8:11, 4:7, 1:3)])
```

## NMF annotations 

```{r message=FALSE}
nmf_GSEA <- readRDS("output/montoro-nmf-gsea.rds")
make_tbl(nmf_GSEA, nnlm_ksort[c(12:30, 5:11, 1:4)])
```

## Topic model annotations

```{r message=FALSE}
tm_GSEA <- readRDS("output/montoro-tm-gsea.rds")
make_tbl(tm_GSEA, tm_ksort[c(16:30, 6:15, 1:5)])
```

