---
title: "Visualizations"
author: "Jason Willwerscheid"
date: "4/25/2022"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(Polychrome)
library(fastTopics)
library(cowplot)
library(ggrepel)
```

## Heatmaps

Main idea: demonstrate the clear difference in sparsity (and interpretability) between vanilla NMF and empirical Bayes NMF.

```{r heatmaps, fig.height=16, fig.width=10, warning=FALSE, message=FALSE}
nnebmf <- readRDS("./output/pijuan-sel-nnebmf-k=40.rds")
nmf <- readRDS("./output/pijuan-sel-nmf-k=40.rds")

fl <- nnebmf$fit
FF <- fl$F.pm
FF <- scale(FF, center = FALSE, scale = apply(FF, 2, max))
FF <- FF[, -1]
colnames(FF) <- paste0("k", 1:ncol(FF))

# Use same cell type order as paper.
cell.type <- sapply(strsplit(rownames(FF), "_"), `[[`, 6)
cell.type <- str_replace(cell.type, "Haematoendothelial", "Haematoendo.")
cell.type <- factor(cell.type, levels = sort(unique(cell.type))[c(
  5, 11, 30, 27, 8, 19, 21, 6, 28, 33, 31, 24, 23, 4, 1, 17, 22, 20, 2, 3, 12:14,
  10, 26, 32, 7, 25, 18, 34, 36, 16, 15, 29, 37,
  9, 35
)])
tomato <- factor(sapply(strsplit(rownames(FF), "_"), `[[`, 4))

# Downsample the number of cells and sort them using tSNE.
set.seed(666)
cell.idx <- numeric(0)
cell.types <- levels(cell.type)
cell.types <- setdiff(cell.types, c("Doublet", "Stripped"))
for (i in 1:length(cell.types)) {
  which.idx <- which(cell.type == cell.types[i])
  # Downsample common cell types.
  if (length(which.idx) > 100) {
    which.idx <- sample(which.idx, 100)
  }
  # Don't include rare cell types.
  if (length(which.idx) > 20) {
    # Sort using tsne.
    tsne.res <- Rtsne::Rtsne(
      FF[which.idx, ],
      dims = 1,
      pca = FALSE,
      normalize = FALSE,
      perplexity = min(100, floor((length(which.idx) - 1) / 3) - 1),
      theta = 0.1,
      max_iter = 1000,
      eta = 200,
      check_duplicates = FALSE
    )$Y[, 1]
    which.idx <- which.idx[order(tsne.res)]
    cell.idx <- c(cell.idx, which.idx)
  }
}

cell.type <- cell.type[cell.idx]
tomato <- tomato[cell.idx]
cell.type <- droplevels(cell.type)

ebmf.FF <- fl$F.pm[cell.idx, order(-fl$pve)]
nmf.FF <- (t(nmf$fit$h) %*% diag(sqrt(nmf$fit$d)))[cell.idx, ]

make.heatmap.tib <- function(FF) {
  tib <- as_tibble(scale(FF, center = FALSE, scale = apply(FF, 2, max))) %>%
    mutate(Cell.type = cell.type, Tomato = tomato) %>%
    arrange(Cell.type, Tomato) %>%
    mutate(Cell.idx = row_number())

  tib <- tib %>%
    pivot_longer(
      -c(Cell.idx, Cell.type, Tomato),
      names_to = "Factor",
      values_to = "Loading",
      values_drop_na = TRUE
    ) %>%
    mutate(Factor = as.numeric(str_extract(Factor, "[0-9]+")))
  
  return(tib)
}

ebmf.tib <- make.heatmap.tib(ebmf.FF)
nmf.tib <- make.heatmap.tib(nmf.FF)
heatmap.tib <- ebmf.tib %>% mutate(Method = "EBMF") %>%
  bind_rows(nmf.tib %>% mutate(Method = "NMF"))

tib <- heatmap.tib %>%
  group_by(Cell.type, Tomato, Cell.idx) %>%
  summarize()
  
cell_type_breaks <- c(1, which(tib$Cell.type[-1] != tib$Cell.type[-nrow(tib)]))
label_pos <- cell_type_breaks / 2 + c(cell_type_breaks[-1], nrow(tib)) / 2
tomato_breaks <- which(tib$Tomato[-1] != tib$Tomato[-nrow(tib)])
tomato_breaks <- setdiff(tomato_breaks, cell_type_breaks)

ggplot(heatmap.tib, aes(x = Factor, y = -Cell.idx, fill = Loading)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "firebrick") +
  labs(y = "") +
  scale_y_continuous(breaks = -label_pos,
                     minor_breaks = NULL,
                     labels = levels(cell.type)) +
  theme_minimal() +
  geom_hline(yintercept = -cell_type_breaks, size = 0.1) +
  geom_hline(yintercept = -tomato_breaks, size = 0.1, linetype = "dashed") +
  facet_wrap(~Method, ncol = 1) +
  theme(legend.position = "none", 
        strip.text = element_text(size = 16))
```

## Structure plots

Main idea: demonstrate that EBMF factors are more "robust" to the choice of $K$ than topics. 

```{r structplots, fig.height=8, fig.width=10, message=FALSE}
tm20 <- readRDS("./output/pijuan-sel-topics-k=20.rds")
tm30 <- readRDS("./output/pijuan-sel-topics-k=30.rds")
tm40 <- readRDS("./output/pijuan-sel-topics-k=40.rds")

ebmf20 <- readRDS("./output/pijuan-sel-nnebmf-k=20.rds")
ebmf30 <- readRDS("./output/pijuan-sel-nnebmf-k=30.rds")
ebmf40 <- readRDS("./output/pijuan-sel-nnebmf-k=40.rds")

subset.idx <- which(cell.type %in% levels(cell.type)[14:20])

grp <- paste0(
  cell.type[subset.idx],
  ifelse(as.logical(tomato[subset.idx]), " (KO)", " (WT)")
)
grp <- factor(grp, levels = sort(unique(grp))[c(4, 3, 9, 8, 1, 2, 5:7)])

get.tm.FF <- function(tm) {
  tm.FF <- t(scale(t(tm$fit$F), center = FALSE, scale = rowSums(tm$fit$F)))[cell.idx, ]
  tm.FF <- tm.FF[subset.idx, ]
  return(tm.FF)
}

get.ebmf.FF <- function(ebmf) {
  ebmf.FF <- (scale(ebmf$fit$F.pm, center = FALSE) %*% diag(sqrt(ebmf$fit$pve)))[cell.idx, order(-ebmf$fit$pve)]
  ebmf.FF <- ebmf.FF[subset.idx, -1]
  return(ebmf.FF)
}

# There are only 31 usable colors, so we need to remove some factors:
ebmf.ref.FF <- get.ebmf.FF(ebmf40)
which.factors <- sort(order(-apply(ebmf.ref.FF, 2, max))[1:31])
ebmf.ref.FF <- ebmf.ref.FF[, which.factors]

tm.ref.FF <- get.tm.FF(tm40)
which.factors <- sort(order(-apply(tm.ref.FF, 2, max))[1:31])
tm.ref.FF <- tm.ref.FF[, which.factors]

# Keep cell order the same across plots:
ref.fit <- list(L = ebmf.ref.FF, F = ebmf.ref.FF)
class(ref.fit) <- "multinom_topic_model_fit"
ebmf.cell.order <- NULL
for (group in levels(grp)) {
  i <- which(grp == group)
  if (length(i) > 0)
    y <- tsne_from_topics(select_loadings(ref.fit, i), dims = 1, verbose = FALSE)
  ebmf.cell.order <- c(ebmf.cell.order, i[order(y)])
}

ref.fit <- list(L = tm.ref.FF, F = tm.ref.FF)
class(ref.fit) <- "multinom_topic_model_fit"
tm.cell.order <- NULL
for (group in levels(grp)) {
  i <- which(grp == group)
  if (length(i) > 0)
    y <- tsne_from_topics(select_loadings(ref.fit, i), dims = 1, verbose = FALSE)
  tm.cell.order <- c(tm.cell.order, i[order(y)])
}

do.struct.plot <- function(LL, ref.FF, cell.order) {
  topic.matches <- rep(0, ncol(LL))
  cormat <- cor(LL, ref.FF)
  for (i in 1:ncol(LL)) {
    j <- row(cormat)[which.max(cormat)]
    k <- col(cormat)[which.max(cormat)]
    topic.matches[j] <- k
    cormat[j, ] <- 0
    cormat[, k] <- 0
  }

  topic.colors <- rep("#808080", ncol(LL))
  for (i in 1:ncol(LL)) {
    if (topic.matches[i] > 0) {
      topic.colors[i] <-  as.character(glasbey.colors(32)[-1])[topic.matches[i]]
    }
  }

  fit <- list(L = LL, F = LL)
  class(fit) <- "multinom_topic_model_fit"

  set.seed(666)
  structure_plot(
    fit,
    grouping = grp,
    colors = topic.colors,
    topics = order(topic.matches),
    loadings_order = cell.order,
    gap = 10
  ) +
    theme(
      legend.position = "none",
      axis.text.y = element_blank(),
    ) +
    labs(y = "")
}

ebmf20p <- do.struct.plot(get.ebmf.FF(ebmf20), ebmf.ref.FF, ebmf.cell.order) +
  labs(title = "EBMF", y = "k = 20") +
  theme(
    axis.text.x = element_blank(),
    axis.title.y = element_text(size = 10),
    plot.title = element_text(size = 16)
  )
ebmf30p <- do.struct.plot(get.ebmf.FF(ebmf30), ebmf.ref.FF, ebmf.cell.order) +
  labs(y = "k = 30") +
  theme(
    axis.text.x = element_blank(),
    axis.title.y = element_text(size = 10)
  )
ebmf40p <- do.struct.plot(get.ebmf.FF(ebmf40), ebmf.ref.FF, ebmf.cell.order) +
  labs(y = "k = 40") +
  theme(axis.title.y = element_text(size = 10))

tm20p <- do.struct.plot(get.tm.FF(tm20), tm.ref.FF, tm.cell.order) +
  labs(title = "Topic Model") +
  theme(
    axis.text.x = element_blank(),
    plot.title = element_text(size = 16)
  )
tm30p <- do.struct.plot(get.tm.FF(tm30), tm.ref.FF, tm.cell.order) +
  theme(axis.text.x = element_blank())
tm40p <- do.struct.plot(get.tm.FF(tm40), tm.ref.FF, tm.cell.order)

plot_grid(
  ebmf20p, ebmf30p, ebmf40p, tm20p, tm30p, tm40p,
  ncol = 2,
  byrow = FALSE,
  rel_heights = c(1.12, 1, 1.65)
)
```

## Volcano plots

Main idea: discuss the difference in interpretation between nonnegative and semi-nonnegative factors. Are the negative loadings interpretable/plausible? We'll need to find a story to tell with this figure.

```{r volcano.code, fig.height=10, fig.width=8, warning=FALSE, message=FALSE}
snnebmf <- readRDS("./output/pijuan-sel-snnebmf-k=40.rds")
exprmean <- log10(readRDS("./data/pijuan-sala-mean-expr.rds")$var.gene.mean.expr)

make.volcano.tib <- function(fl, k, exprmean) {
  tib <- tibble(
    pm = fl$L.pm[, k] / max(abs(fl$L.pm[, k])),
    z = abs(fl$L.pm[, k]) / pmax(sqrt(.Machine$double.eps), fl$L.psd[, k]),
    SYMBOL = rownames(fl$L.pm),
    exprmean = exprmean
  ) %>%
    mutate(z.thresh = ifelse(abs(pm) > .1, z, 0)) %>%
    mutate(SYMBOL = ifelse(
      z.thresh > sort(z.thresh, decreasing = TRUE)[9] |
        (pm > sort(pm, decreasing = TRUE)[9]) |
        (pm < 0 & pm < sort(pm, decreasing = FALSE)[6]), 
      SYMBOL, 
      ""
    )) 
}

do.volcano.plot <- function(nn.fl, snn.fl, kset, exprmean, k.labels) {
  tib <- tibble()
  for (i in 1:length(kset)) {
    k <- kset[i]
    snn.k <- which.max(abs(cor(snn.fl$F.pm[, -1], nn.fl$F.pm[, k]))) + 1
    tib <- tib %>%
      bind_rows(
        make.volcano.tib(nn.fl, k, exprmean) %>% 
          mutate(k = k.labels[i], Type = "Nonnegative")
      ) %>%
      bind_rows(
        make.volcano.tib(snn.fl, snn.k, exprmean) %>% 
          mutate(k = k.labels[i], Type = "Semi-nonnegative")
      )
  }
  tib$k <- factor(tib$k, levels = k.labels)
  
  plt <- ggplot(tib, aes(x = pm, y = z, label = SYMBOL, color = exprmean)) +
    geom_point() +
    scale_color_gradient2(low = "deepskyblue", mid = "gold", high = "orangered",
                          na.value = "gainsboro",
                          midpoint = mean(range(exprmean))) +
    scale_y_continuous(trans = "log1p", breaks = 10^(1:floor(log10(max(tib$z))))) +
    geom_text_repel(color = "darkgray",size = 2.25, fontface = "italic",
                    segment.color = "darkgray", segment.size = 0.25,
                    min.segment.length = 0, na.rm = TRUE, max.overlaps = 20) +
    theme_minimal() +
    labs(
      x = "Gene Loading (posterior mean)",
      y = "|z-score|",
      color = "Mean Expression (log10)",
    ) +
    theme(legend.position = "bottom") +
    facet_grid(
      rows = vars(k), 
      cols = vars(Type),
      scales = "free_x",
      space = "free_x"
    ) +
    theme(
      strip.text.x = element_text(size = 16),
      strip.text.y = element_text(size = 10, angle = 0),
      panel.spacing = unit(2, "lines")
    )

  return(plt)
}

kset <- order(-nnebmf$fit$pve)[c(2, 7, 24)]
k.labels <- c("Eryth (2)", "Other WT (7)", "Epi WT (24)")
do.volcano.plot(nnebmf$fit, snnebmf$fit, kset, exprmean, k.labels)
```

